Characteristic polynomial of p-adic matrices

### 2017-01-05 charpoly/charpoly.tex
Minor improvements.

\begin{prop} 
Algorithm 1 computes $H$ and $P$ realizing an approximate Hessenberg form of $M.$
$P$ is exact over finite extensions of $\mathbb{Q}_p$ and $k((X))$, and the computation is in $O(n^3)$ operations in $K$ at precision the maximum precision of a coeffient in $M.$
\end{prop}
\begin{proof}
Let us assume that $K$ is a finite extensions of $\mathbb{Q}_p$ or $k((X)).$
Inside the nested \textbf{for} loop, if we want to eliminate $p^{u_y} \varepsilon_y+O(p^{n_y}$ with pivot $p^{u_x} \varepsilon_x+O(p^{n_x},$
with the $\varepsilon$'s being units,
the coefficient of the transvection matrix is the lift (in $\mathbb{Z}, $  $k[X]$ or adequate extension) of $p^{u_y-u_x} \varepsilon_y \varepsilon_x^{-1} \mod \pi^{\min (n_x,n_y)}.$
Exactness follows directly. Over other fields, we can not lift, but the computations are still valid.
The rest is clear.
\end{proof}

@@

\begin{prop}
Let $cond(M)$ be the valuation of the last invariant factor of $M.$
If $dM$ is a flat precision of $O(\pi^N)$ on $M$ then $M^{-1}$
can be computed at precision $O(\pi^{N-2cond(M)})$ by a \textbf{SNF} computation
and this lower-bound is optimal
at least when $N$ is large.
\end{prop}
\begin{proof}
The smallest valuation of a coefficient of $M^{-1}$ is $-cond(M).$
It is $-2cond(M)$ for $M^{-2}$ 2 and it is then clear that $-2cond(M)$
can be obtained as the valuation of a coefficient of $du(M) \cdot dM$
and the smallest that can be achieved this way for $dM$ in a precision lattice
of flat precision. Hence the optimality of the bound given, at least when 
$N$ is large, thanks to \todo{Reference for precision lemma}.

Now, the computation of an SNF form over a CDVF was described in \todo{reference these Vaccon}.
From $M$ known at flat precision $O(\pi^N),$ we can obtain an exact $\Delta$ and $P,Q$ 
known at precision at least $O(\pi^{N-cond(M)})$ with coefficients in \todo{integral ring}
and determinant in \todo{integer ring invertible} realizing an SNF form of $M.$
There is no loss in precision when computing $P^{-1}$ and $Q^{-1}.$
Now, computing $\Delta^{-1}$ exactly, its smallest coefficient being  of valuation $-cond(M),$
we obtain $M^{-1}=Q^{-1} \Delta^{-1} P^{-1}$ known at precision at least $O(\pi^{N-2cond(M)}),$
which concludes the proof.
\end{proof}

@@

Firstly, we need a lemma to relate comatrices for
similar matrices:

@@

The first ingredients are reciprocal polynomials.
We extend their definition to matrices of polynomials.
\begin{deftn}
Let $A \in M_n(K[X]).$ We denote by $A^{rec}$ the matrix
whose coefficients are the reciprocal polynomials of the
coefficients of $A.$ 
\end{deftn}
We then have the following result :
\begin{lem}
Let $M \in M_n(K).$ Then:
\begin{eqnarray*}
com(1-XM)^{rec}=&com(X-M), \\
(\chi_M \times 1)^{rec}=&(1-XM) com(1-XM).\\
\end{eqnarray*}
\end{lem}

@@

\begin{theo}
Let $H \in M_n(O_K)$ be an approximate Hessenberg matrix.
Then, using Algorithm 2, one can compute $\com (X -H)$ in 
$\softO (n^3)$ operations in $O_K$ at the precision given by $H.$
\end{theo}
\begin{proof}
Firstly, the operations of the lines 2 and 3 are in $\softO (n^3)$ operations
in $O_K$ at the precision given by $H.$
Indeed, since $H$ is an approximate matrix, when we use $U_{i,i}$ as pivot,
on its columns, only $U_{i+1,i}$ can be a non zero-approximate coefficient.
As a consequence, when performing this column-pivoting, only two rows ($i$ and
$i+1$) lead to operations in $O_K [[K]]$ other than checking precision.
Hence, line 3 costs $\softO (n^2)$ for the computation of $U.$
Following line 1, the computation of $Q^{-1}$ is done by operations on rows, starting from the identity matrix.
Coefficient for the transvection is the opposite of the one defined by pivoting
and the indices for the receiving and giving rows are swapped compared to the operations on columns.
Thus, we can show that it requires no additional operation in $O_K[[X]].$ $Q^{-1}$ is just filled as
an upper-triangular matrix.
This proves that the resulting cost for lines 2 and 3 is indeed in $\softO (n^3)$ operations.

@@

\begin{prop} \label{prop:optimal_flat}
If $M \in M_n(O_K)$ is known at flat precision $dM$ (high enough), then 
we can compute the optimal jagged precision on $\chi (M)$
in $\softO (n^3)$ operations in $O_K$ at the initial flat precision.
\end{prop}
\begin{proof}
Computing $M=PHP^{-1}$ an approximate Hessenberg form can be done in
$O(n^3)$ with no division.
Computing $\com (X-H)$ can be done in $\softO (n^3)$ with no division.
Then, by Lemma \ref{lem:comatrix_of_similar}, $tr(\com (X-M) \cdot dM)=tr(P\com (X-H)P^{-1} \cdot dM).$
thus, \[tr(\com (X-M) \cdot dM)=tr(\com (X-H) P^{-1}  dM P). \]
$dM$ being flat, so is $P^{-1}  dM P$ with same precision, 
and the jagged precision on $\chi (M)$ can then directly be read on $\com (X-H).$
\end{proof}

\subsection{$\com (X -M)$ and jagged precision}

@@

When this time $dM$ is jagged, the trick of the proof of Proposition
\ref{prop:optimal_flat} can not work anymore.
Indeed, the computation of $P^{-1}  dM P$ and then of $\com (X-H) P^{-1}  dM P$
are not easy as $P^{-1}  dM P$ is not jagged anymore.
Fortunately, a factorization result of $\com (X-M)$ will be enough
to conclude that optimal precision can still be read in $\softO (n^3)$
operations in $O_K.$

@@

Now, let us take $U=col(A,1)$ the first column of $A$ and
$V=A_{1,1}^{-1} row(A,1) \mod \chi (M)$ (as $A_{1,1}$ is
invertible $\mod \chi (M).$
Then thanks to \eqref{eqn:2-minors-com}, $U$ and $V$ satisfy the desired equality. 
\end{proof}

@@

\begin{theo}
Algorithm 3 compute 
$\com (X-M) \mod (\chi(M))$ in average complexity $\softO (n^3)$
operations in $O_K$ at initial precision.
\end{theo}
\begin{proof}
As we have already seen, completing Step 1 and 2 is in $\softO (n^3).$
Multiplying by $T$ or $S$ or their inverse corresponds
to $n$ operations on rows or columns over a matrix with coefficients
in $O_K[X]$ of degree at most $n.$
Thus, it is in $\softO (n^3).$
Step 5 and 6 are also in $\softO (n^3).$
All that is to prove is that the set of $P$ and $S$ to avoid
is of positive dimension.
As any the proof of \ref{prop:factor_com}, we can work modulo $X-\lambda$
for $\lambda$ a root of $\chi (M)$ (in an algebraic closuer)
and then apply Chinese Remainder Theorem.
The goal of the Step $2$ is to ensure the first row of $B$ contains an
invertible modulo $\chi (M).$
Since $A(\lambda)$ is of rank one, the $\mu_i$'s have to avoid an
affine hyperplane so that $row(B,1) \mod (X-\lambda)$ is a non-zero vector.
Hence for $row(B,1) \mod \chi (M)$ to contain an invertible coefficient,
a finite union of affine hyperplane is to avoid.
Similarily, the goal of Step 3 is to put an invertible coefficient (modulo
$\chi(M)$) on $C_{1,1},$ and again, only a finite union of affine
hyperplane is to avoid.
Hence, the set that the $\mu_i$'s have to avoid is a finite union
of hyperplane, and hence, is of positive dimension.
Thus, almost any $\mu_i$ lead to a matrix $C$ passing the test
on Step 4, which concludes the proof.
\end{proof}

@@

\begin{rem}
As a consequence of the previous remark, once the optimal jagged
precision is known, it is possible to possible lift arbitrarily
the precision on $M$ and then to perform
a computation of the characteristic polynomial
from an Hessenberg form as in \ref{rem:char_pol_from_hessenberg},
in $O(n^3).$
By truncating afterward at the optimal jagged precision,
we can achieve by this way this optimal precision.
\end{rem}

>>>>

\begin{prop} 
Algorithm 1 computes $H$ and $P$ realizing an approximate Hessenberg form of $M.$
$P$ is exact over finite extensions of $\mathbb{Q}_p$ and $k((X))$, and the computation is in $O(n^3)$ operations in $K$ at precision the maximum precision of a coeffient in $M.$
\end{prop}
\begin{proof}
Let us assume that $K$ is a finite extensions of $\mathbb{Q}_p$ or $k((X)).$
Inside the nested \textbf{for} loop, if we want to eliminate $p^{u_y} \varepsilon_y+O(p^{n_y})$ with pivot $p^{u_x} \varepsilon_x+O(p^{n_x}),$
with the $\varepsilon$'s being units,
the coefficient of the transvection matrix is the lift (in $\mathbb{Z}, $  $k[X]$ or adequate extension) of $p^{u_y-u_x} \varepsilon_y \varepsilon_x^{-1} \mod \pi^{\min (n_x,n_y)}.$
Exactness follows directly. Over other fields, we can not lift, but the computations are still valid.
The rest is clear.
\end{proof}

@@

\begin{prop}
Let $cond(M)$ be the valuation of the last invariant factor of $M.$
If $dM$ is a flat precision of $O(\pi^N)$ on $M$ then $M^{-1}$
can be computed at precision $O(\pi^{N-2cond(M)})$ by a \textbf{SNF} computation
and this lower-bound is optimal
at least when $N$ is large.
\end{prop}
\begin{proof}
The smallest valuation of a coefficient of $M^{-1}$ is $-cond(M).$
It is $-2cond(M)$ for $M^{-2}$ and it is then clear that $-2cond(M)$
can be obtained as the valuation of a coefficient of $du(M) \cdot dM$
and the smallest that can be achieved this way for $dM$ in a precision lattice
of flat precision. Hence the optimality of the bound given, at least when 
$N$ is large, thanks to \todo{Reference for precision lemma}.

Now, the computation of an SNF form over a CDVF was described in \todo{reference these Vaccon}.
From $M$ known at flat precision $O(\pi^N),$ we can obtain an exact $\Delta$ and $P,Q$ 
known at precision at least $O(\pi^{N-cond(M)})$ with coefficients in $O_K$
and determinant in $O_K^*$ realizing an SNF form of $M.$
There is no loss in precision when computing $P^{-1}$ and $Q^{-1}.$
Now, computing $\Delta^{-1}$ exactly, its smallest coefficient being  of valuation $-cond(M),$
we obtain $M^{-1}=Q^{-1} \Delta^{-1} P^{-1}$ known at precision at least $O(\pi^{N-2cond(M)}),$
which concludes the proof.
\end{proof}

@@

Firstly, we need a lemma to relate comatrices of
similar matrices:

@@

The second ingredient we need is the reciprocal polynomial.
We extend its definition to matrices of polynomials.
\begin{deftn}
Let $A \in M_n(K[X]).$ We denote by $A^{rec}$ the matrix
whose coefficients are the reciprocal polynomials of the
coefficients of $A.$ 
\end{deftn}
We then have the following result :
\begin{lem}
Let $M \in M_n(K).$ Then:
\begin{eqnarray*}
com(1-XM)^{rec}=&com(X-M), \\
(\chi_M \times 1)^{rec}=&(1-XM) com(1-XM).\\
\end{eqnarray*}
\end{lem}

@@

\begin{theo}
Let $H \in M_n(O_K)$ be an approximate Hessenberg matrix.
Then, using Algorithm 2, one can compute $\com (X -H)$ in 
$\softO (n^3)$ operations in $O_K$ at the precision given by $H.$
\end{theo}
\begin{proof}
Firstly, the operations of the lines 2 and 3 are in $\softO (n^3)$ operations
in $O_K$ at the precision given by $H.$
Indeed, since $H$ is an approximate matrix, when we use $U_{i,i}$ as pivot:
on its columns, only $U_{i+1,i}$ can be a non zero-approximate coefficient.
As a consequence, when performing this column-pivoting, only two rows ($i$ and
$i+1$) lead to operations in $O_K [[K]]$ other than checking precision.
Hence, line 3 costs $\softO (n^2)$ for the computation of $U.$
Following line 1, the computation of $Q^{-1}$ is done by operations on rows, starting from the identity matrix.
Coefficient for the transvection is the opposite of the one defined by pivoting
and the indices for the receiving and giving rows are swapped compared to the operations on columns.
Thus, we can show that it requires no additional operation in $O_K[[X]].$ $Q^{-1}$ is just filled as
an upper-triangular matrix.
This proves that the resulting cost for lines 2 and 3 is indeed in $\softO (n^3)$ operations.

@@

\begin{prop} \label{prop:optimal_flat}
If $M \in M_n(O_K)$ is known at flat precision $dM$ (high enough), then 
we can compute the optimal jagged precision on $\chi (M)$
in $\softO (n^3)$ operations in $O_K$ at the initial flat precision.
\end{prop}
\begin{proof}
Computing $M=PHP^{-1},$ an approximate Hessenberg form, can be done in
$O(n^3)$ with no division.
Computing $\com (X-H)$ can be done in $\softO (n^3)$ with no division.
Then, by Lemma \ref{lem:comatrix_of_similar}, $tr(\com (X-M) \cdot dM)=tr(P\com (X-H)P^{-1} \cdot dM).$
thus, \[tr(\com (X-M) \cdot dM)=tr(\com (X-H) P^{-1}  dM P). \]
$dM$ being flat, so is $P^{-1}  dM P$ with same precision, 
and the jagged precision on $\chi (M)$ can then directly be read on $\com (X-H).$
\end{proof}

%\subsection{$\com (X -M)$ and jagged precision}

@@

When this time $dM$ is jagged, the trick of the proof of Proposition
\ref{prop:optimal_flat} can not work anymore.
Indeed, the computation of $P^{-1}  dM P$ and then of $\com (X-H) P^{-1}  dM P$
are not easy, as $P^{-1}  dM P$ is not jagged anymore.
Fortunately, a factorization result of $\com (X-M)$ will be enough
to conclude that optimal precision can still be read in $\softO (n^3)$
operations in $O_K.$

@@

Now, let us take $U=col(A,1)$ the first column of $A$ and
$V=A_{1,1}^{-1} row(A,1) \mod \chi (M)$ (as $A_{1,1}$ is
invertible $\mod \chi (M)).$
Then thanks to \eqref{eqn:2-minors-com}, $U$ and $V$ satisfy the desired equality. 
\end{proof}

@@

\begin{theo}
Algorithm 3 compute 
$\com (X-M) \mod (\chi(M))$ in average complexity $\softO (n^3)$
operations in $O_K$ at initial precision.
\end{theo}
\begin{proof}
As we have already seen, completing Step 1 and 2 is in $\softO (n^3).$
Multiplying by $T$ or $S$ or their inverse corresponds
to $n$ operations on rows or columns over a matrix with coefficients
in $O_K[X]$ of degree at most $n.$
Thus, it is in $\softO (n^3).$
Step 5 and 6 are also in $\softO (n^3).$
All that is to prove is that the set of $P$ and $S$ to avoid
is of dimension at most $n-1.$
As in the proof of \ref{prop:factor_com}, we can work modulo $X-\lambda$
for $\lambda$ a root of $\chi (M)$ (in an algebraic closure)
and then apply Chinese Remainder Theorem.
The goal of the Step $2$ is to ensure the first row of $B$ contains an
invertible modulo $\chi (M).$
Since $A(\lambda)$ is of rank one, the $\mu_i$'s have to avoid an
affine hyperplane so that $row(B,1) \mod (X-\lambda)$ is a non-zero vector.
Hence for $row(B,1) \mod \chi (M)$ to contain an invertible coefficient,
a finite union of affine hyperplane is to avoid.
Similarily, the goal of Step 3 is to put an invertible coefficient (modulo
$\chi(M)$) on $C_{1,1},$ and again, only a finite union of affine
hyperplane is to avoid.
Hence, the set that the $\mu_i$'s have to avoid is a finite union
of hyperplane, and hence, is of dimension at most $n-1.$
Thus, almost any $\mu_i$ lead to a matrix $C$ passing the test
on Step 4, which concludes the proof.
\end{proof}

@@

\begin{rem}
As a consequence of the previous remark, once the optimal jagged
precision is known, it is possible to lift arbitrarily
the precision on $M$ and then to perform
a computation of the characteristic polynomial
from an Hessenberg form as in \ref{rem:char_pol_from_hessenberg},
in $O(n^3).$
By truncating afterward at the optimal jagged precision,
we can achieve by this way this optimal precision.
\end{rem}
